Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-09 paper/paper.tex
Add a little more foreshadowing in Section 3

Our system is predicative because the interpretation of the $[[i]]$th universe
is dependent only on universes strictly below $[[i]]$,
which have already been defined.
This restriction ensures that the relation is well defined; otherwise, the
definition of $[[InterpR i A S]]$ would not be well founded, and
$[[Interp I i A S]]$ could call $[[I]]$ on universe levels greater than
or equal to $[[i]]$, which are yet to be defined.
%% SCW: removing the constraing wouldn't result in any system at all
% Removing the ordering constraint would result in a
% system where one can encode Girard's
% paradox~\citep{girard-thesis}.

@@

\subsection{Properties of the Logical Relation}
In the rest of this section, we develop the theory of our logical relation
with the goal of showing four key facts: irrelevance
(Lemma~\ref{lemma:logrelcoherence}), functionality
(Lemma~\ref{lemma:logreldeter}), cumulativity
(Lemma~\ref{lemma:logrelcumulativity}), and the backward closure property
(Lemma~\ref{lemma:logrelbackclos}).  For the majority of the properties that
we prove in this section, we do not need any information about the
parameterized function $[[I]]$.  Each property about $[[InterpR i A S]]$
follows as a corollary of a property about $[[Interp I i A S]]$ with no or few
assumptions imposed on $[[I]]$. As a result, we state most of our lemmas in
terms of $[[Interp I i A S]]$ without duplicating them in terms of
$[[InterpR i A S]]$.

>>>>

Our system is predicative because the interpretation of the $[[i]]$th universe
is dependent only on universes strictly below $[[i]]$.
% which have already been defined.
This restriction ensures that the relation is well defined; otherwise, the
definition of $[[InterpR i A S]]$ would not be well founded, and
$[[Interp I i A S]]$ could call $[[I]]$ on universe levels greater than
or equal to $[[i]]$, which are yet to be defined.
%% SCW: removing the constraing wouldn't result in any system at all
% Removing the ordering constraint would result in a
% system where one can encode Girard's
% paradox~\citep{girard-thesis}.

@@

\subsection{Properties of the Logical Relation}

In the rest of this section, we develop the theory of our logical relation
with the goal of showing four key properties: irrelevance
(Lemma~\ref{lemma:logrelcoherence}), functionality
(Lemma~\ref{lemma:logreldeter}), cumulativity
(Lemma~\ref{lemma:logrelcumulativity}), and the backward closure property
(Lemma~\ref{lemma:logrelbackclos}).
These properties semantically correspond to ones that hold for syntactic typing,
and allow us to prove that syntactic typing implies semantic typing, which we
define in the next section.

In particular, irrelevance states that convertible
types have the same interpretation, which is used to interpret \rref{T-Conv}.
Functionality with cumulativity shows that all interpretations of a type are
the same across all universe levels, which allows us to treat a type that
appears twice in a judgment uniformly, such as the function domain type in
\rref{T-App}. Finally, backward closure ensures that anything that reduces to
a term in the interpretation of a type is itself in the interpretation, which
we use to handle $\beta$-reductions in the cases for introduction forms.

For the majority of the properties that we prove in this section, we need no
information about the parameterized function $[[I]]$.  Each property about
$[[InterpR i A S]]$ follows as a corollary of a property about
$[[Interp I i A S]]$ with no or few assumptions imposed on $[[I]]$. As a result,
we state most of our lemmas in terms of $[[Interp I i A S]]$ without duplicating
them in terms of $[[InterpR i A S]]$.
